---
title       : Team Stats Explorer for Major League Baseball
subtitle    : Coursera Developing Data Products Course Project
author      : Marc Hidalgo
job         : 
framework   : io2012
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Description of Application

- This application allows users to view statistics for Major League Baseball (MLB) teams over the period 1969 - 2000

- Data are viewable by MLB franchises that were in existence during that period (for reference, see [MLB Franchise History](https://en.wikipedia.org/wiki/Timeline_of_Major_League_Baseball)).

- Resultant plots are color coded by general team colors to allow easier inspection of the data. 

- Data for the entire period is plotted for the selected teams(s), but, in some cases (e.g. Blue Jays, Rockies, Marlins), the team came into existence sometime during 1969 - 2000. In these cases, data will only be plotted for the year the franchise existed.

---

```{r, echo=FALSE}
library(ggplot2)
library(data.table)
library(rCharts)
library(reshape2)
library(markdown)
library(UsingR)

MLBattend$color <- "black"
MLBattend$color[MLBattend$franchise == "ANA"] <- "red"
MLBattend$franchise <- gsub("ANA","Angels",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "CAL"] <- "red"
MLBattend$franchise <- gsub("CAL","Angels",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "ARI"] <- "darkred"
MLBattend$franchise <- gsub("ARI","Diamond Backs",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "ATL"] <- "red1"
MLBattend$franchise <- gsub("ATL","Braves",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "BAL"] <- "darkorange"
MLBattend$franchise <- gsub("BAL","Orioles",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "BOS"] <- "red2"
MLBattend$franchise <- gsub("BOS","Red Sox",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "CHIA"] <- "slategray4"
MLBattend$franchise <- gsub("CHIA","White Sox",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "CHIN"] <- "blue4"
MLBattend$franchise <- gsub("CHIN","Cubs",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "CIN"] <- "red2"
MLBattend$franchise <- gsub("CIN","Reds",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "CLE"] <- "darkblue"
MLBattend$franchise <- gsub("CLE","Indians",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "COL"] <- "blueviolet"
MLBattend$franchise <- gsub("COL","Rockies",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "DET"] <- "midnightblue"
MLBattend$franchise <- gsub("DET","Tigers",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "FL"] <- "darkturquoise"
MLBattend$franchise <- gsub("FL","Marlins",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "HOU"] <- "red3"
MLBattend$franchise <- gsub("HOU","Astros",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "KC"] <- "royalblue4"
MLBattend$franchise <- gsub("KC","Royals",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "MILA"] <- "navyblue"
MLBattend$franchise <- gsub("MILA","Brewers",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "MILN"] <- "navyblue"
MLBattend$franchise <- gsub("MILN","Brewers",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "LA"] <- "dodgerblue"
MLBattend$franchise <- gsub("LA","Dodgers",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "MIN"] <- "midnightblue"
MLBattend$franchise <- gsub("MIN","Twins",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "MON"] <- "mediumblue"
MLBattend$franchise <- gsub("MON","Expos",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "NYA"] <- "navy"
MLBattend$franchise <- gsub("NYA","Yankees",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "NYN"] <- "blue2"
MLBattend$franchise <- gsub("NYN","Mets",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "OAK"] <- "springgreen4"
MLBattend$franchise <- gsub("OAK","A's",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "PHI"] <- "indianred4"
MLBattend$franchise <- gsub("PHI","Phillies",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "PIT"] <- "black"
MLBattend$franchise <- gsub("PIT","Pirates",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "SD"] <- "navyblue"
MLBattend$franchise <- gsub("SD","Padres",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "SEA"] <- "darkslategray"
MLBattend$franchise <- gsub("SEA","Mariners",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "SF"] <- "darkorange3"
MLBattend$franchise <- gsub("SF","Giants",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "STL"] <- "red1"
MLBattend$franchise <- gsub("STL","Cardinals",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "TB"] <- "turquoise4"
MLBattend$franchise <- gsub("TB","Rays",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "TEX"] <- "firebrick"
MLBattend$franchise <- gsub("TEX","Rangers",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "WAS"] <- "firebrick4"
MLBattend$franchise <- gsub("WAS","Rangers",MLBattend$franchise, fixed=TRUE)
MLBattend$color[MLBattend$franchise == "TOR"] <- "blue3"
MLBattend$franchise <- gsub("TOR","Blue Jays",MLBattend$franchise, fixed=TRUE)

#sort the team list to be in alphabetical order
teamList <- sort(unique(MLBattend$franchise))

#adjust the year range from 69 - 00 to be from 1969 - 2000
MLBattend$year <- 1900 + MLBattend$year
MLBattend$year[MLBattend$year == 1900] <- 2000
MLBattend$attendance <- MLBattend$attendance/100000

#create MLB and League averages by Year for a reference plot
MLByears <- sort(unique(MLBattend$year))
MLBfranchise <- "MLB"
MLBleague <- "NA"
MLBdivision <- "NA"
MLBavgAttend <- mean(MLBattend$attendance[MLBattend$year==MLByears[1]])
MLBavgScored <- mean(MLBattend$runs.scored[MLBattend$year==MLByears[1]])
MLBavgAllowed <- mean(MLBattend$runs.allowed[MLBattend$year==MLByears[1]])
MLBavgWins <- mean(MLBattend$wins[MLBattend$year==MLByears[1]])
MLBavgLosses <- mean(MLBattend$losses[MLBattend$year==MLByears[1]])
MLBavgBehind <- mean(MLBattend$games.behind[MLBattend$year==MLByears[1]])
for(i in 2:length(MLByears))
{
  MLBfranchise <- c(MLBfranchise,"MLB")
  MLBleague <- c(MLBleague,"NA")
  MLBdivision <- c(MLBdivision,"NA")
  MLBavgAttend <- c(MLBavgAttend,mean(MLBattend$attendance[MLBattend$year==MLByears[i]]))
  MLBavgScored <- c(MLBavgScored,mean(MLBattend$runs.scored[MLBattend$year==MLByears[i]]))
  MLBavgAllowed <- c(MLBavgAllowed,mean(MLBattend$runs.allowed[MLBattend$year==MLByears[i]]))
  MLBavgWins <- c(MLBavgWins,mean(MLBattend$wins[MLBattend$year==MLByears[i]]))
  MLBavgLosses <- c(MLBavgLosses,mean(MLBattend$losses[MLBattend$year==MLByears[i]]))
  MLBavgBehind <- c(MLBavgBehind,mean(MLBattend$games.behind[MLBattend$year==MLByears[i]]))
}
MLBavg <- data.frame(franchise = MLBfranchise, league = MLBleague, division = MLBdivision,
                     year = MLByears, attendance = MLBavgAttend, 
                     runs.scored = MLBavgScored, runs.allowed = MLBavgAllowed,
                     wins = MLBavgWins, losses = MLBavgLosses,
                     games.behind = MLBavgBehind)

```

Users may select any of the following teams

```{r, echo=FALSE}
teamList
```


Available statistics include: 

- Yearly attendance

- Total runs scored

- Total runs allowed

- Wins

---

## Example Plots

```{r, echo=FALSE}
      par(mfrow=c(2,1))
      plot(MLBavg$year,MLBavg$attendance,col='gray', type='l', lwd=3, lty=2,
           xlim=c(1969,2000),ylim=c(0,50), main="Attendance by Year",
           ylab="attendance (100K)", xlab="year")
      #plot data for the user selected team(s)
      lines(MLBattend$year[MLBattend$franchise == "Angels"],
            MLBattend$attendance[MLBattend$franchise == "Angels"],
            col=MLBattend$color[MLBattend$franchise == "Angels"], lwd=4)
      lines(MLBattend$year[MLBattend$franchise == "Cubs"],
            MLBattend$attendance[MLBattend$franchise == "Cubs"],
            col=MLBattend$color[MLBattend$franchise == "Cubs"], lwd=4)
      lines(MLBattend$year[MLBattend$franchise == "A's"],
            MLBattend$attendance[MLBattend$franchise == "A's"],
            col=MLBattend$color[MLBattend$franchise == "A's"], lwd=4)
      plot(MLBavg$year,MLBavg$attendance, type='l',col='white',xlab=' ',ylab=' ', bg='white', 
           col.axis='white', col.lab='white', col.main='white', fg='white',
           xlim=c(1969,2000),ylim=c(0,50))
      #add legend for the MLB average line centered at the top
      lines(c(1974,1982),c(45,45),lwd=3,col='gray',lty=2)
      text(1983,45,"MLB Average (1969 - 2000)", pos=4)
      lines(c(1970,1974),c(35,35),lwd=4,
            col=MLBattend$color[MLBattend$franchise == "Angels"])
      text(1975,35,"Angels", pos=4)
      lines(c(1982,1986),c(35,35),lwd=4,
            col=MLBattend$color[MLBattend$franchise == "Cubs"])
      text(1987,35,"Cubs", pos=4)
      column <- 1
      lines(c(1994,1998),c(35,35),lwd=4,
            col=MLBattend$color[MLBattend$franchise == "A's"])
      text(1999,35,"A's", pos=4)
```

--- 

## More Information

- Data used for the application are associated with the UsingR library of the R programming language [UsingR](https://cran.r-project.org/web/packages/UsingR/) available from [Comprehensive R Archive Network](https://cran.r-project.org/). 

- Source code is available on  [GitHub](https://github.com/juanmarc/developing-data-products-project).

- A demo of the application can be found [here](https://juanmarc.shinyapps.io/developing-data-products-project/).
